---
title: "Bayesian Hierarchical GAMM for mCPR"
subtitle: "DHS country trajectories, 2026 growth-headroom investment staging"
author: "Analytics Team"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
options(mc.cores = max(1, parallel::detectCores() - 1))
```

# Executive Summary

This report fits a **Bayesian hierarchical GAMM** to DHS mCPR data by country and year, then uses **2026 model-based predictions** to classify countries into investment stages.

Core outputs:

- Country-level fitted mCPR trajectories with uncertainty
- Growth (
\(\frac{dp}{dt}\)
) and acceleration (
\(\frac{d^2p}{dt^2}\)
) patterns
- Implicit country saturation proxy and 2026 headroom
- Investment-stage classification and priority ranking for 2026 decisions

# Modeling Objective

For country \(i\) in survey year \(t\), let:

- \(y_{it}\): number of women using modern contraception
- \(n_{it}\): number of women observed
- \(p_{it}\): modern method prevalence probability

Likelihood:

\[
y_{it} \sim \text{Binomial}(n_{it}, p_{it})
\]

Hierarchical additive predictor:

\[
\text{logit}(p_{it}) = \beta_0 + b_{0i} + f_0(t) + f_i(t)
\]

where:

- \(f_0(t)\): global smooth over time
- \(f_i(t)\): country-specific smooth deviation (factor-smooth)
- \(b_{0i} \sim \mathcal{N}(0, \sigma_b)\): country random intercept

This formulation keeps the outcome in **binomial space** (no pre-logit transformed response), which is the correct setup for `brms` with `family = binomial(link = "logit")`.

# Libraries

```{r libraries}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(brms)
library(posterior)
library(DT)
library(scales)
```

# Data Preparation

```{r load-and-prepare-data}
# Expected columns: country, year, mCPR (0-1), n_women
mcpr_raw <- read.csv("mcpr_data.csv", stringsAsFactors = FALSE)

required_cols <- c("country", "year", "mCPR", "n_women")
stopifnot(all(required_cols %in% names(mcpr_raw)))

mcpr_dat <- mcpr_raw %>%
  transmute(
    country = factor(country),
    year = as.numeric(year),
    mCPR = as.numeric(mCPR),
    n = as.integer(round(n_women))
  ) %>%
  mutate(
    mCPR = pmin(pmax(mCPR, 1e-6), 1 - 1e-6),
    y = as.integer(round(mCPR * n))
  ) %>%
  filter(!is.na(country), !is.na(year), !is.na(y), !is.na(n), n > 0, y >= 0, y <= n) %>%
  arrange(country, year)

YEAR_REF <- 2000
DECISION_YEAR <- 2026
GRID_STEP <- 0.5
HORIZON_YEAR <- max(DECISION_YEAR + 10, max(mcpr_dat$year, na.rm = TRUE) + 6)

mcpr_dat <- mcpr_dat %>%
  mutate(
    year_c = year - YEAR_REF,
    obs_prop = y / n
  )

cat("Rows:", nrow(mcpr_dat), "\n")
cat("Countries:", n_distinct(mcpr_dat$country), "\n")
cat("Year range:", min(mcpr_dat$year), "to", max(mcpr_dat$year), "\n")
cat("Decision year:", DECISION_YEAR, "\n")
```

```{r data-coverage-table}
mcpr_dat %>%
  group_by(country) %>%
  summarise(
    surveys = n(),
    first_year = min(year),
    last_year = max(year),
    min_mCPR = min(obs_prop),
    max_mCPR = max(obs_prop),
    .groups = "drop"
  ) %>%
  arrange(country) %>%
  DT::datatable(
    caption = "Country survey coverage",
    options = list(pageLength = 12, scrollX = TRUE),
    rownames = FALSE
  )
```

# Initial Global Shape and Dynamics (Descriptive)

This section gives an initial (non-model) view of global shape and acceleration/deceleration using country-year means.

```{r descriptive-global-shape}
global_obs <- mcpr_dat %>%
  group_by(year) %>%
  summarise(
    mean_mcpr = mean(obs_prop),
    .groups = "drop"
  ) %>%
  arrange(year) %>%
  mutate(
    d1 = (lead(mean_mcpr) - lag(mean_mcpr)) / (lead(year) - lag(year)),
    d2 = (lead(mean_mcpr) - 2 * mean_mcpr + lag(mean_mcpr)) / (((lead(year) - lag(year)) / 2)^2)
  )

ggplot(global_obs, aes(x = year, y = mean_mcpr)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.8, linetype = "dashed") +
  theme_minimal(base_size = 12) +
  labs(
    title = "Descriptive global mCPR shape (country-average)",
    x = "Year",
    y = "Mean observed mCPR"
  )
```

```{r descriptive-global-dynamics}
global_dyn <- global_obs %>%
  select(year, d1, d2) %>%
  pivot_longer(cols = c(d1, d2), names_to = "metric", values_to = "value") %>%
  mutate(
    metric = recode(metric, d1 = "First derivative (growth)", d2 = "Second derivative (acceleration)")
  )

ggplot(global_dyn, aes(x = year, y = value)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5, alpha = 0.8) +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Descriptive global growth and acceleration signals",
    x = "Year",
    y = "Value"
  )
```

# Bayesian Hierarchical GAMM

## Formula

```{r model-formula}
bgam_formula <- bf(
  y | trials(n) ~ 1 +
    (1 | country) +
    s(year_c, k = 10) +
    s(year_c, by = country, bs = "fs", k = 6)
)

bgam_formula
```

## Priors

We center the intercept prior at the weighted global observed mCPR:

\[
\mu_0 = \text{logit}\left(\frac{\sum y}{\sum n}\right)
\]

Priors:

- \(\beta_0 \sim \mathcal{N}(\mu_0, 1.5)\)
- \(\sigma_b \sim \mathcal{N}^+(0, 1)\)
- spline SDs \(\sim t^+_{\nu=3}(0, 1.5)\)

```{r model-priors}
p_bar <- with(mcpr_dat, sum(y) / sum(n))
p_bar <- pmin(pmax(p_bar, 1e-4), 1 - 1e-4)
mu0 <- qlogis(p_bar)

bgam_priors <- c(
  set_prior(sprintf("normal(%0.8f, 1.5)", mu0), class = "Intercept"),
  prior(normal(0, 1), class = "sd", group = "country"),
  prior(student_t(3, 0, 1.5), class = "sds")
)

cat("Weighted p_bar =", round(p_bar, 4), "| logit center mu0 =", round(mu0, 3), "\n")
```

## Fit

```{r fit-model, cache=TRUE}
fit_path <- "fit_bgam_mcpr_2026.rds"

if (file.exists(fit_path)) {
  fit_bgam <- readRDS(fit_path)
} else {
  fit_bgam <- brm(
    formula = bgam_formula,
    data = mcpr_dat,
    family = binomial(link = "logit"),
    prior = bgam_priors,
    chains = 4,
    iter = 2500,
    warmup = 1250,
    seed = 123,
    control = list(adapt_delta = 0.995, max_treedepth = 15),
    save_pars = save_pars(all = TRUE)
  )
  saveRDS(fit_bgam, fit_path)
}
```

```{r fit-summary}
summary(fit_bgam)
```

# Diagnostics

```{r diagnostics-plot}
plot(fit_bgam)
```

```{r posterior-predictive-checks}
pp_check(fit_bgam, type = "bars", ndraws = 200)
pp_check(fit_bgam, type = "error_hist", ndraws = 200)
```

# Fitted Values and Residual Checks

```{r fitted-values}
ep_obs <- posterior_epred(fit_bgam, ndraws = 500)
p_hat <- colMeans(ep_obs)

mcpr_fit <- mcpr_dat %>%
  mutate(
    p_hat = p_hat,
    resid_raw = obs_prop - p_hat,
    p_clip = pmin(pmax(p_hat, 1e-6), 1 - 1e-6),
    resid_pearson = (y - n * p_hat) / sqrt(n * p_clip * (1 - p_clip))
  )

summary(mcpr_fit$p_hat)
summary(mcpr_fit$resid_pearson)
```

```{r residual-plots}
p1 <- ggplot(mcpr_fit, aes(x = year, y = resid_pearson)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.8) +
  theme_minimal(base_size = 12) +
  labs(title = "Pearson residuals vs year", x = "Year", y = "Pearson residual")

p2 <- ggplot(mcpr_fit, aes(x = p_hat, y = resid_pearson)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.8) +
  theme_minimal(base_size = 12) +
  labs(title = "Pearson residuals vs fitted", x = "Fitted mCPR", y = "Pearson residual")

p1
p2
```

# Prediction Grids and Derivative Utilities

```{r helper-functions}
finite_diff <- function(mat, h) {
  # mat: draws x time
  n_t <- ncol(mat)
  d1 <- matrix(NA_real_, nrow(mat), n_t)
  d2 <- matrix(NA_real_, nrow(mat), n_t)
  if (n_t >= 3) {
    d1[, 2:(n_t - 1)] <- (mat[, 3:n_t, drop = FALSE] - mat[, 1:(n_t - 2), drop = FALSE]) / (2 * h)
    d2[, 2:(n_t - 1)] <- (mat[, 3:n_t, drop = FALSE] - 2 * mat[, 2:(n_t - 1), drop = FALSE] + mat[, 1:(n_t - 2), drop = FALSE]) / (h^2)
  }
  list(d1 = d1, d2 = d2)
}

summarize_draw_matrix <- function(mat, probs = c(0.05, 0.5, 0.95)) {
  tibble(
    q05 = apply(mat, 2, quantile, probs = probs[1], na.rm = TRUE),
    q50 = apply(mat, 2, quantile, probs = probs[2], na.rm = TRUE),
    q95 = apply(mat, 2, quantile, probs = probs[3], na.rm = TRUE)
  )
}

rescale_safe <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (!all(is.finite(rng)) || diff(rng) == 0) {
    return(rep(0, length(x)))
  }
  (x - rng[1]) / diff(rng)
}
```

# Global Model-Based Shape, Growth, and Acceleration

```{r global-grid-and-predictions, cache=TRUE}
year_seq_global <- seq(min(mcpr_dat$year), DECISION_YEAR, by = GRID_STEP)

global_grid <- tidyr::crossing(
  country = levels(mcpr_dat$country),
  year = year_seq_global
) %>%
  mutate(
    year_c = year - YEAR_REF,
    n = 1L
  )

ep_global <- posterior_epred(fit_bgam, newdata = global_grid, ndraws = 500)

idx_by_year <- split(seq_len(nrow(global_grid)), global_grid$year)

global_draws <- sapply(idx_by_year, function(idx) {
  rowMeans(ep_global[, idx, drop = FALSE])
})

global_draws <- as.matrix(global_draws)
colnames(global_draws) <- names(idx_by_year)

global_der <- finite_diff(global_draws, h = GRID_STEP)

global_summ <- bind_cols(
  tibble(year = as.numeric(colnames(global_draws))),
  summarize_draw_matrix(global_draws) %>% rename(p05 = q05, p50 = q50, p95 = q95),
  summarize_draw_matrix(global_der$d1) %>% rename(dp05 = q05, dp50 = q50, dp95 = q95),
  summarize_draw_matrix(global_der$d2) %>% rename(d205 = q05, d250 = q50, d295 = q95)
)
```

```{r global-shape-plot}
ggplot(global_summ, aes(x = year, y = p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.25) +
  geom_line(linewidth = 1) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Global model-based mCPR trajectory (country-average)",
    subtitle = "Median and 90% credible interval",
    x = "Year",
    y = "mCPR"
  )
```

```{r global-derivative-plot}
global_der_long <- global_summ %>%
  select(year, dp05, dp50, dp95, d205, d250, d295) %>%
  mutate(
    growth = dp50,
    growth_lo = dp05,
    growth_hi = dp95,
    accel = d250,
    accel_lo = d205,
    accel_hi = d295
  )

p_growth <- ggplot(global_der_long, aes(x = year, y = growth)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_ribbon(aes(ymin = growth_lo, ymax = growth_hi), alpha = 0.25) +
  geom_line(linewidth = 0.9) +
  theme_minimal(base_size = 12) +
  labs(title = "Global growth (dp/dt)", x = "Year", y = "mCPR per year")

p_accel <- ggplot(global_der_long, aes(x = year, y = accel)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_ribbon(aes(ymin = accel_lo, ymax = accel_hi), alpha = 0.25) +
  geom_line(linewidth = 0.9) +
  theme_minimal(base_size = 12) +
  labs(title = "Global acceleration (d2p/dt2)", x = "Year", y = "mCPR per year^2")

p_growth
p_accel
```

# Country Curves, Growth, and Acceleration

```{r country-grid-predictions, cache=TRUE}
year_seq_country <- seq(min(mcpr_dat$year), HORIZON_YEAR, by = GRID_STEP)

country_grid <- tidyr::crossing(
  country = levels(mcpr_dat$country),
  year = year_seq_country
) %>%
  mutate(
    year_c = year - YEAR_REF,
    n = 1L
  ) %>%
  arrange(country, year)

ep_country <- posterior_epred(fit_bgam, newdata = country_grid, ndraws = 500)

idx_by_country <- split(seq_len(nrow(country_grid)), country_grid$country)

country_curve <- map_dfr(names(idx_by_country), function(cty) {
  idx <- idx_by_country[[cty]]
  yr <- country_grid$year[idx]
  mat <- ep_country[, idx, drop = FALSE]
  der <- finite_diff(mat, h = GRID_STEP)

  bind_cols(
    tibble(country = cty, year = yr),
    summarize_draw_matrix(mat) %>% rename(p05 = q05, p50 = q50, p95 = q95),
    summarize_draw_matrix(der$d1) %>% rename(dp05 = q05, dp50 = q50, dp95 = q95),
    summarize_draw_matrix(der$d2) %>% rename(d205 = q05, d250 = q50, d295 = q95)
  )
}) %>%
  mutate(country = factor(country, levels = levels(mcpr_dat$country)))
```

```{r country-fits-facet, fig.width=14, fig.height=18}
plot_data_fit <- country_curve %>% filter(year <= DECISION_YEAR)

ggplot(plot_data_fit, aes(x = year, y = p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.2) +
  geom_line(linewidth = 0.7) +
  geom_point(data = mcpr_dat, aes(x = year, y = obs_prop), inherit.aes = FALSE, alpha = 0.5, size = 1.2) +
  facet_wrap(~ country, scales = "free_x", ncol = 5) +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(size = 8, face = "bold")) +
  labs(
    title = "Country-level Bayesian GAMM fitted trajectories",
    subtitle = "Observed points + posterior median and 90% CrI",
    x = "Year",
    y = "mCPR"
  )
```

```{r country-growth-accel-facet, fig.width=14, fig.height=18}
plot_data_dyn <- country_curve %>% filter(year <= DECISION_YEAR)

p_ctry_growth <- ggplot(plot_data_dyn, aes(x = year, y = dp50)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_ribbon(aes(ymin = dp05, ymax = dp95), alpha = 0.2) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ country, scales = "free_x", ncol = 5) +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(size = 8, face = "bold")) +
  labs(title = "Country growth dp/dt", x = "Year", y = "mCPR per year")

p_ctry_accel <- ggplot(plot_data_dyn, aes(x = year, y = d250)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_ribbon(aes(ymin = d205, ymax = d295), alpha = 0.2) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ country, scales = "free_x", ncol = 5) +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(size = 8, face = "bold")) +
  labs(title = "Country acceleration d2p/dt2", x = "Year", y = "mCPR per year^2")

p_ctry_growth
p_ctry_accel
```

# 2026 Decision Metrics

We classify investment stage using **model predictions at 2026**.

Definitions by country \(i\):

- Current level: \(\hat{p}_{i,2026}\)
- Current growth: \(\widehat{dp/dt}_{i,2026}\)
- Implicit ceiling proxy: \(\hat{K}_i = Q_{0.95}\left(\hat{p}_i(t),\ t \in [2026, \text{horizon}]\right)\)
- Headroom: \(\hat{h}_i = \max(\hat{K}_i - \hat{p}_{i,2026}, 0)\)

```{r decision-metrics-2026}
country_2026 <- country_curve %>%
  group_by(country) %>%
  filter(abs(year - DECISION_YEAR) == min(abs(year - DECISION_YEAR))) %>%
  slice(1) %>%
  ungroup() %>%
  transmute(
    country,
    year_decision = DECISION_YEAR,
    p2026 = p50,
    p2026_lo = p05,
    p2026_hi = p95,
    growth2026 = dp50,
    growth2026_lo = dp05,
    growth2026_hi = dp95,
    accel2026 = d250,
    accel2026_lo = d205,
    accel2026_hi = d295
  )

country_ceiling <- country_curve %>%
  filter(year >= DECISION_YEAR) %>%
  group_by(country) %>%
  summarise(
    K90 = quantile(p50, 0.90, na.rm = TRUE),
    K95 = quantile(p50, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

country_status <- country_2026 %>%
  left_join(country_ceiling, by = "country") %>%
  mutate(
    headroom = pmax(K95 - p2026, 0),
    rel_saturation = p2026 / pmax(K95, 1e-6)
  )

summary(country_status$headroom)
summary(country_status$growth2026)
```

# Investment Stage Classification (2026)

Classification is based on joint growth-headroom position:

- High growth + high headroom: strongest scale-up potential
- Mixed profile: targeted strategy and constraint-specific support
- Low growth + low headroom: maintenance/equity focus

```{r stage-classification}
growth_cut <- quantile(country_status$growth2026, 2/3, na.rm = TRUE)
headroom_cut <- quantile(country_status$headroom, 2/3, na.rm = TRUE)

country_stage <- country_status %>%
  mutate(
    growth_norm = rescale_safe(pmax(growth2026, 0)),
    headroom_norm = rescale_safe(headroom),
    accel_bonus = ifelse(accel2026 > 0, 0.10, 0),
    priority_score = 100 * pmin(1, pmax(0, 0.60 * growth_norm + 0.40 * headroom_norm + accel_bonus)),
    stage = case_when(
      growth2026 >= growth_cut & headroom >= headroom_cut ~ "Stage 1: Accelerate investment",
      growth2026 < growth_cut & headroom < headroom_cut ~ "Stage 3: Sustain and equity",
      TRUE ~ "Stage 2: Targeted scale-up"
    ),
    stage_rank = case_when(
      stage == "Stage 1: Accelerate investment" ~ 1L,
      stage == "Stage 2: Targeted scale-up" ~ 2L,
      TRUE ~ 3L
    ),
    accel_label = case_when(
      accel2026 > 0 ~ "Accelerating",
      accel2026 < 0 ~ "Decelerating",
      TRUE ~ "Flat"
    )
  ) %>%
  arrange(stage_rank, desc(priority_score))

cat("Growth cut (67th pct):", round(growth_cut, 4), "\n")
cat("Headroom cut (67th pct):", round(headroom_cut, 4), "\n")
```

# 2026 Investment Ranking Table

```{r ranking-table}
investment_table <- country_stage %>%
  transmute(
    country,
    stage,
    stage_rank,
    priority_score,
    p2026,
    p2026_lo,
    p2026_hi,
    growth2026,
    growth2026_lo,
    growth2026_hi,
    accel2026,
    accel_label,
    K90,
    K95,
    headroom,
    rel_saturation
  ) %>%
  arrange(stage_rank, desc(priority_score))

DT::datatable(
  investment_table %>% mutate(across(where(is.numeric), ~ round(.x, 4))),
  caption = "Country investment staging and priority (decision year = 2026)",
  filter = "top",
  options = list(pageLength = 25, scrollX = TRUE),
  rownames = FALSE
)
```

# Investment Visualizations

```{r stage-count-plot}
stage_counts <- country_stage %>%
  count(stage, stage_rank) %>%
  arrange(stage_rank)

ggplot(stage_counts, aes(x = reorder(stage, stage_rank), y = n, fill = stage)) +
  geom_col(alpha = 0.9, show.legend = FALSE) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  labs(
    title = "Countries by investment stage (2026)",
    x = NULL,
    y = "Number of countries"
  )
```

```{r headroom-growth-scatter}
ggplot(country_stage, aes(x = growth2026, y = headroom, color = stage, size = priority_score)) +
  geom_vline(xintercept = growth_cut, linetype = "dashed", alpha = 0.7) +
  geom_hline(yintercept = headroom_cut, linetype = "dashed", alpha = 0.7) +
  geom_point(alpha = 0.85) +
  theme_minimal(base_size = 12) +
  scale_size_continuous(range = c(2, 8)) +
  labs(
    title = "2026 growth vs headroom investment map",
    subtitle = "Top-right quadrant = strongest scale-up profile",
    x = "Growth at 2026 (dp/dt)",
    y = "Headroom (K95 - p2026)",
    color = NULL,
    size = "Priority"
  ) +
  theme(legend.position = "bottom")
```

```{r selected-country-trajectories}
selected_countries <- country_stage %>%
  group_by(stage) %>%
  arrange(desc(priority_score), .by_group = TRUE) %>%
  slice_head(n = 3) %>%
  ungroup() %>%
  pull(country)

selected_curve <- country_curve %>%
  filter(country %in% selected_countries, year <= DECISION_YEAR)

selected_stage <- country_stage %>%
  select(country, stage, priority_score)

selected_obs <- mcpr_dat %>%
  filter(country %in% selected_countries)

ggplot(selected_curve, aes(x = year, y = p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.2) +
  geom_line(linewidth = 0.9) +
  geom_point(data = selected_obs, aes(x = year, y = obs_prop), inherit.aes = FALSE, alpha = 0.6, size = 1.5) +
  geom_vline(xintercept = DECISION_YEAR, linetype = "dotted") +
  facet_wrap(~ country, scales = "free_y") +
  theme_minimal(base_size = 11) +
  labs(
    title = "Illustrative countries by investment stage",
    subtitle = "Top-priority examples per stage",
    x = "Year",
    y = "mCPR"
  )
```

# Interpretation Notes

- Stage 1 countries combine **high current growth** and **substantial headroom** in 2026, making them strongest candidates for marginal investment returns.
- Stage 2 countries have one strong dimension (growth or headroom), indicating targeted rather than broad scale-up.
- Stage 3 countries are relatively saturated and/or slow-growing by 2026, shifting emphasis toward maintenance, quality, and equity.

# Technical Appendix

```{r session-info}
cat("R:", R.version.string, "\n")
cat("brms:", as.character(packageVersion("brms")), "\n")
cat("posterior:", as.character(packageVersion("posterior")), "\n")
```
