---
title: "get dhs"
output: html_document
date: "2026-02-08"
---

## 0: Set up packages and one-time rdhs config file

```{r}



library(rdhs)
library(dplyr)
library(tibble)
library(purrr)
library(survey)
library(haven)

## One-time per machine: set rdhs config (then comment out)
# cache_dir <- "~/dhs_data_cache"
# if (!dir.exists(cache_dir)) dir.create(cache_dir, recursive = TRUE)
#
# set_rdhs_config(
#   email       = "gracia.sierra@utexas.edu",
#   project     = "Contraceptive transition over time a cross country comparison",
#   cache_path  = cache_dir,
#   config_path = "~/.rdhs.json",
#   global      = TRUE
# )

```

## 1. Identify IR datasets (women’s Individual Recode), 1985+
- include common DHS delivery formats (Stata/SPSS/SAS/Flat)
- optional file exclusions for persistent bad files only

`ir_meta` defines the files we'll obtain through the API

```{r}

allowed_formats <- c("Stata", "SPSS", "SAS", "Flat")
file_exclusions <- character(0)  # add filenames only if a file is persistently broken

ir_meta <- dhs_datasets(fileType = "IR") %>%
  mutate(Year = as.numeric(SurveyYear)) %>%
  filter(SurveyType == "DHS",
    Year >= 1985,
    grepl(paste(allowed_formats, collapse = "|"), FileFormat, ignore.case = TRUE),
    !FileName %in% file_exclusions
  ) %>%
  distinct(SurveyId, FileName, .keep_all = TRUE) %>%
  arrange(CountryName, Year)


cat("IR datasets to process:", nrow(ir_meta), "\n")
```

## 2. Design based mCPR computation - per survey

```{r}

load_cached_ir_dataset <- function(paths) {
  # We only need the design and contraceptive-use fields for mCPR extraction.
  needed_vars <- c("v012", "v313", "v005", "v021", "v022", "v023")
  needed_vars <- unique(c(needed_vars, toupper(needed_vars)))

  # rdhs usually gives us a cached .rds file, which is fastest to load.
  # Some large surveys in the local cache are corrupted, so we fall back to
  # the original Stata/SPSS files if the .rds cannot be read.
  read_direct_file <- function(path) {
    ext <- tolower(tools::file_ext(path))
    if (ext == "rds") {
      return(as.data.frame(readRDS(path)))
    }
    if (ext == "dta") {
      return(as.data.frame(haven::read_dta(
        path,
        col_select = tidyselect::any_of(needed_vars)
      )))
    }
    if (ext == "sav") {
      return(as.data.frame(haven::read_sav(
        path,
        col_select = tidyselect::any_of(needed_vars)
      )))
    }
    stop("Unsupported file extension: ", ext)
  }

  candidate_paths <- unique(paths[file.exists(paths)])
  if (length(candidate_paths) == 0) {
    stop("No cached dataset paths were found")
  }

  preferred_paths <- c(
    candidate_paths[grepl("\\.rds$", candidate_paths, ignore.case = TRUE)],
    candidate_paths[grepl("\\.(dta|sav)$", candidate_paths, ignore.case = TRUE)]
  )

  for (path in preferred_paths) {
    dat <- tryCatch(read_direct_file(path), error = function(e) NULL)
    if (!is.null(dat)) {
      return(dat)
    }
  }

  zip_paths <- candidate_paths[grepl("\\.zip$", candidate_paths, ignore.case = TRUE)]
  for (zip_path in zip_paths) {
    zip_listing <- tryCatch(unzip(zip_path, list = TRUE), error = function(e) NULL)
    if (is.null(zip_listing) || nrow(zip_listing) == 0) next

    zip_members <- zip_listing$Name
    data_member <- zip_members[grepl("\\.(dta|sav)$", zip_members, ignore.case = TRUE)][1]
    if (is.na(data_member)) next

    extract_dir <- tempfile("dhs_extract_")
    dir.create(extract_dir, recursive = TRUE, showWarnings = FALSE)
    extracted_path <- tryCatch(
      unzip(zip_path, files = data_member, exdir = extract_dir),
      error = function(e) NULL
    )
    if (is.null(extracted_path) || length(extracted_path) == 0) next

    dat <- tryCatch(read_direct_file(extracted_path[[1]]), error = function(e) NULL)
    if (!is.null(dat)) {
      return(dat)
    }
  }

  stop("Unable to read any cached dataset artifact for this survey")
}

calc_mCPR_from_path <- function(paths, meta_row) {
  # 1. Load the dataset from the cached rdhs files, with a raw-file fallback.
  dat <- load_cached_ir_dataset(paths)
  dat <- as.data.frame(dat)

  # Case-insensitive variable lookup helper
  vn <- names(dat)
  vn_low <- tolower(vn)
  get_var <- function(name) {
    idx <- which(vn_low == tolower(name))[1]
    if (is.na(idx)) return(NULL)
    dat[[idx]]
  }

  # 2. Restrict to women 15–49 (v012)
  v012 <- get_var("v012")
  if (!is.null(v012)) {
    keep <- !is.na(v012) & v012 >= 15 & v012 <= 49
    dat <- dat[keep, , drop = FALSE]
    vn <- names(dat)
    vn_low <- tolower(vn)
    get_var <- function(name) {
      idx <- which(vn_low == tolower(name))[1]
      if (is.na(idx)) return(NULL)
      dat[[idx]]
    }
  }

  if (is.null(nrow(dat)) || nrow(dat) == 0) {
    return(tibble(
      country     = meta_row$CountryName,
      survey_id   = meta_row$SurveyId,
      year        = as.numeric(meta_row$Year),
      n_rows      = 0L,
      n_women     = NA_real_,
      y_weighted  = NA_real_,
      mCPR        = NA_real_,
      mCPR_pct    = NA_real_,
      se_mCPR     = NA_real_,
      se_mCPR_pct = NA_real_,
      error_msg   = NA_character_
    ))
  }

  # 3. Modern use via v313.
  # Early DHS rounds do not use a universal numeric code for "modern method".
  # We therefore read the survey-specific value labels and detect the code that
  # is actually labelled "modern method", falling back to the common 0/1/2/3
  # scheme only when labels are absent.
  v313 <- get_var("v313")
  if (is.null(v313)) {
    return(tibble(
      country     = meta_row$CountryName,
      survey_id   = meta_row$SurveyId,
      year        = as.numeric(meta_row$Year),
      n_rows      = nrow(dat),
      n_women     = NA_real_,
      y_weighted  = NA_real_,
      mCPR        = NA_real_,
      mCPR_pct    = NA_real_,
      se_mCPR     = NA_real_,
      se_mCPR_pct = NA_real_,
      error_msg   = "Missing v313"
    ))
  }
  v313_labels <- attr(v313, "labels")
  modern_codes <- numeric(0)
  nonmodern_codes <- numeric(0)

  if (!is.null(v313_labels) && !is.null(names(v313_labels))) {
    label_names <- tolower(trimws(names(v313_labels)))
    modern_codes <- unname(v313_labels[grepl("\\bmodern\\b", label_names)])
    nonmodern_codes <- unique(unname(v313_labels[
      grepl("no method|not using|folkloric|traditional", label_names)
    ]))
  }

  if (length(modern_codes) == 0) modern_codes <- 3
  if (length(nonmodern_codes) == 0) nonmodern_codes <- c(0, 1, 2)

  modern_use <- dplyr::case_when(
    v313 %in% modern_codes ~ 1,
    v313 %in% nonmodern_codes ~ 0,
    TRUE ~ NA_real_
  )

  # 4. Weight via v005
  v005 <- get_var("v005")
  if (is.null(v005)) {
    return(tibble(
      country     = meta_row$CountryName,
      survey_id   = meta_row$SurveyId,
      year        = as.numeric(meta_row$Year),
      n_rows      = nrow(dat),
      n_women     = NA_real_,
      y_weighted  = NA_real_,
      mCPR        = NA_real_,
      mCPR_pct    = NA_real_,
      se_mCPR     = NA_real_,
      se_mCPR_pct = NA_real_,
      error_msg   = "Missing v005"
    ))
  }
  weight <- v005 / 1e6

  # Drop NAs
  valid <- !is.na(modern_use) & !is.na(weight)
  modern_use <- modern_use[valid]
  weight     <- weight[valid]
  dat_valid  <- dat[valid, , drop = FALSE]

  if (length(modern_use) == 0) {
    return(tibble(
      country     = meta_row$CountryName,
      survey_id   = meta_row$SurveyId,
      year        = as.numeric(meta_row$Year),
      n_rows      = nrow(dat),
      n_women     = NA_real_,
      y_weighted  = NA_real_,
      mCPR        = NA_real_,
      mCPR_pct    = NA_real_,
      se_mCPR     = NA_real_,
      se_mCPR_pct = NA_real_,
      error_msg   = "No valid rows after filtering"
    ))
  }

  # 5. Weighted mCPR
  n_women <- sum(weight, na.rm = TRUE)
  y       <- sum(modern_use * weight, na.rm = TRUE)
  p_hat   <- if (n_women > 0) y / n_women else NA_real_

  # 6. Design-based SE using DHS design variables when available.
  # Fallback: Kish effective sample size approximation.
  v021 <- get_var("v021")
  v022 <- get_var("v022")
  v023 <- get_var("v023")
  strata <- if (!is.null(v022)) v022 else v023

  se_hat <- NA_real_
  if (!is.null(v021) && length(v021) == nrow(dat) && length(weight) == nrow(dat_valid)) {
    v021 <- v021[valid]
    strata <- if (!is.null(strata) && length(strata) == nrow(dat)) strata[valid] else NULL

    design_df <- tibble(
      modern_use = modern_use,
      weight = weight,
      cluster = v021,
      strata = if (is.null(strata)) NA_real_ else strata
    )

    design_df <- design_df %>%
      filter(!is.na(modern_use), !is.na(weight), !is.na(cluster), weight > 0)

    if (nrow(design_df) > 0) {
      svy_obj <- tryCatch({
        if (all(is.na(design_df$strata))) {
          svydesign(ids = ~cluster, weights = ~weight, data = design_df, nest = TRUE)
        } else {
          design_df <- design_df %>% filter(!is.na(strata))
          svydesign(ids = ~cluster, strata = ~strata, weights = ~weight, data = design_df, nest = TRUE)
        }
      }, error = function(e) NULL)

      if (!is.null(svy_obj)) {
        svy_est <- tryCatch(svymean(~modern_use, svy_obj, na.rm = TRUE), error = function(e) NULL)
        if (!is.null(svy_est)) {
          se_hat <- as.numeric(SE(svy_est))[1]
        }
      }
    }
  }

  if (is.na(se_hat)) {
    w <- weight
    n_eff <- (sum(w)^2) / sum(w^2)
    se_hat <- if (!is.na(p_hat) && n_eff > 0) sqrt(p_hat * (1 - p_hat) / n_eff) else NA_real_
  }

  tibble(
    country     = meta_row$CountryName,
    survey_id   = meta_row$SurveyId,
    year        = as.numeric(meta_row$Year),
    n_rows      = nrow(dat),
    n_women     = n_women,
    y_weighted  = y,
    mCPR        = p_hat,
    mCPR_pct    = 100 * p_hat,
    se_mCPR     = se_hat,
    se_mCPR_pct = 100 * se_hat,
    error_msg   = NA_character_
  )
}



```

## 3. STREAMING loop: survey → mCPR + SE, no microdata stored

```{r}
results_list <- vector("list", nrow(ir_meta))

for (i in seq_len(nrow(ir_meta))) {
  meta_row <- ir_meta[i, ]
  cat("Processing", i, "/", nrow(ir_meta), ":",
      meta_row$CountryName, meta_row$SurveyYear, "…\n")

  results_list[[i]] <- tryCatch({
    paths <- get_datasets(meta_row$FileName)
    calc_mCPR_from_path(paths, meta_row)
  }, error = function(e) {
    tibble(
      country     = meta_row$CountryName,
      survey_id   = meta_row$SurveyId,
      year        = as.numeric(meta_row$Year),
      n_rows      = NA_real_,
      n_women     = NA_real_,
      y_weighted  = NA_real_,
      mCPR        = NA_real_,
      mCPR_pct    = NA_real_,
      se_mCPR     = NA_real_,
      se_mCPR_pct = NA_real_,
      error_msg   = as.character(e$message)
    )
  })

  gc()
}

mCPR_survey <- dplyr::bind_rows(results_list)
mCPR_errors <- mCPR_survey %>% filter(!is.na(error_msg))

# Aggregate to one row per country-year for modeling.
mCPR_design_all <- mCPR_survey %>%
  group_by(country, year) %>%
  summarise(
    survey_id = paste(sort(unique(survey_id)), collapse = ";"),
    n_surveys_year = n(),
    n_rows = sum(n_rows, na.rm = TRUE),
    n_women = sum(n_women, na.rm = TRUE),
    y_weighted = sum(y_weighted, na.rm = TRUE),
    mCPR = ifelse(n_women > 0, y_weighted / n_women, NA_real_),
    mCPR_pct = 100 * mCPR,
    se_mCPR = if (n() == 1) {
      dplyr::first(se_mCPR)
    } else if (all(is.na(se_mCPR))) {
      NA_real_
    } else {
      max(se_mCPR, na.rm = TRUE)
    },
    se_mCPR_pct = 100 * se_mCPR,
    .groups = "drop"
  ) %>%
  arrange(country, year)

mCPR_design <- mCPR_design_all %>%
  filter(!is.na(mCPR), !is.na(n_women), n_women > 0)

summary(mCPR_design$mCPR_pct)


cat("Design-based mCPR rows:", nrow(mCPR_design), "\n")

```


## 4. Sanity checks + validation

```{r}

# Surveys
cat("Surveys in ir_meta:   ", nrow(ir_meta), "\n")
cat("Survey rows in mCPR_survey:", nrow(mCPR_survey), "\n")
cat("Country-year rows (all):", nrow(mCPR_design_all), "\n")
cat("Country-year rows (model-ready):", nrow(mCPR_design), "\n")

# Countries
cat("Countries in metadata:", ir_meta %>% distinct(CountryName) %>% nrow(), "\n")
cat("Countries in mCPR_design:", mCPR_design %>% distinct(country) %>% nrow(), "\n")
cat("Rows with extraction errors:", nrow(mCPR_errors), "\n")
cat("Country-years with >1 source survey:", sum(mCPR_design_all$n_surveys_year > 1), "\n")

# N and mCPR distribution
cat("\nSummary of n_women (weighted):\n")
print(summary(mCPR_design$n_women))

cat("\nSummary of mCPR (percent):\n")
print(summary(mCPR_design$mCPR_pct))

cat("\nSummary of SE(mCPR) (percent points):\n")
print(summary(mCPR_design$se_mCPR_pct))

# Peek at a few rows
mCPR_design %>%
  arrange(country, year) %>%
  select(country, year, survey_id, n_women, mCPR_pct, se_mCPR_pct) %>%
  head(30)
```

## 5. Save output for downstream modeling

```{r}
write.csv(
  mCPR_design %>% arrange(country, year),
  "mcpr_data.csv",
  row.names = FALSE
)

write.csv(
  mCPR_errors %>% arrange(country, year),
  "mcpr_data_extraction_errors.csv",
  row.names = FALSE
)
```

```{r}

```



```{r}

```

```{r}

```

```{r}

```



```{r}

```

```{r}

```

